Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study in India

### 2025-07-22 manuscript.Rmd
Incorporate the additional online material in the manuscript

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

@@

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS^®^ and PTC training. We collected data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline. In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].

@@

Data was collected over a four-month period. A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care. The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records. We followed up admitted patients for their complications and other in-hospital outcome measures. Patients who were not admitted or who were discharged before the end of the study were followed up telephonically for mortality outcomes and quality of life outcomes.

@@

We used the `r version$version.string` Statistical Software for all analyses [@R_2024]. We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014]. Initially, we planned to analyse feasibility outcomes both on an overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results. Quantitative variables are summarised as median and interquartile range. Qualitative variables are presented as absolute numbers and percentages. Additional analyses performed according to the original protocol are described in the Supplementary Material S2 and their results shown in Supplementary Material S3 and Supplementary Tables S1-12.

@@

We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`. The period of data collection and the number of patients included per month per cluster is shown in Table \@ref(tab:patients-per-month). The standard care arm included `r n.control` patients, the ATLS^®^ arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients. We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.

```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month."}
knitr::include_graphics("patients-per-month.png")
```

The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. These prognostic factors differed between the trial arms. A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.

@@

The percentage of patients who consented to follow up was `r p.consent`% and the lost to follow up rate was `r rate.lost.to.follow.up`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the maximum amount of missing data were in the cost of treatment, reported in Supplementary Tables S1-12.

@@

We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete. To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.

@@

There are some important limitations of this pilot study and therefore additional lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. The full scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial. Finally, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.

>>>>

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988; @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016; @kamau_impact_2024; @kumar_evaluating_2025], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

@@

We piloted a three-armed cluster randomised controlled trial [@GerdinWärnberg2022]. There were a standard care arm and two intervention arms, ATLS^®^ and PTC training. We planned to collect data for four months in all three arms, first during a one month observation phase and then during a three month intervention phase (or continued observation in the standard care arm). The actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters. We included the one month lead to be able to evaluate the feasibility of comparing patient outcomes both as absolute differences between the intervention phases, and as differences in change from baseline. In the published protocol, we also aimed to estimate probable effect sizes and other measures needed for the sample size calculations of a full-scale trial [@GerdinWärnberg2022], but we revised this aim in the light of current guidance on the conduct and reporting of pilot trials [@teresi_guidelines_2022].

@@

The planned data collection period was four months, but the actual data collection period varied across clusters depending on the timing of the training, to ensure a minimum of three months of data collection after the training in the intervention clusters. A research officer collected data on all patients who presented on the days and shifts when participating residents were assigned to trauma care. The research officers observed care and interviewed residents and patients, and also extracted data from the hospital records. We followed up admitted patients for their complications and other in-hospital outcome measures. Patients who were not admitted or who were discharged before the end of the study were followed up telephonically for mortality outcomes and quality of life outcomes.

@@

We used the `r version$version.string` Statistical Software for all analyses [@R_2024]. We analysed all data, including the feasibility outcomes, using descriptive statistics and did not perform any formal hypothesis tests [@Lee2014]. Initially, we planned to analyse feasibility outcomes both on an overall and individual cluster levels, but the sample sizes in individual clusters were too small to generate meaningful results. Quantitative variables are summarised as median and interquartile range. Qualitative variables are presented as absolute numbers and percentages. Additional analyses performed according to the original protocol are available as additional online material [@warnberg_additional_2025].

@@

We included `r n.patients` trauma patients from `r number_to_text(n.centres)` clusters between `r start.date` and `r end.date`. The period of data collection and the number of patients included per month per cluster is shown in Figure \@ref(fig:patients-per-month-figure). Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively. The standard care arm included `r n.control` patients, the ATLS^®^ arm included `r n.atls` patients, and the PTC arm included `r n.ptc` patients. We trained a total of  `r number_to_text(n.residents)` residents, `r number_to_text(n.atls.residents)` in ATLS^®^, and  `r number_to_text(n.ptc.residents)` in PTC.

```{r patients-per-month-figure, echo = FALSE, results = "asis", fig.cap = "Number of patients included per cluster per month. Due to an error in the data uploading process, data was only available for one and three months for two clusters respectively."}
knitr::include_graphics("patients-per-month.png")
```

The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Extended patient sample characteristics are shown in Supplementary Table S2. Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. These prognostic factors differed between the trial arms. A total of `r m30d` patients died within 30 days after arrival to the emergency department, and `r inhosp` patients died in hospital.

@@

The percentage of patients who consented to follow up was `r p.consent`% and the lost to follow up rate was `r rate.lost.to.follow.up`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`%, with details for selected variables shown in Table \@ref(tab:sample-characteristics) and in Supplementary Table S2. The variables with the maximum amount of missing data were in the cost of treatment, also reported in Supplementary Tables S2.

@@

We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete. To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.

@@

There are some important limitations of this pilot study and therefore additional lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. The full scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial. Fifth, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes. Finally, because of a data uploading error, there was only limited data available from two clusters. At the time of data collection, network issues and other technical issues were a challenge in some of the clusters, and this can be mitigated in the full scale trial by using a more robust data collection system with local offline backups.
